---
title: "BWAS-RESI R1 - Data preparation for global and regional brain measures data"
author: "Kaidi Kang"
date: "7/25/2023"
output: 
  html_document:
    toc: true
    number_sections: true
    use_bookdown: yes
    code_folding: hide
    toc_float: 
      collapsed: false
editor_options: 
  chunk_output_type: inline
---

```{r}
library(table1)
library(e1071) # for calculating moments
library(pracma) # for the nth root
library(readr)
library(dplyr)
library(magrittr)
library(splines)
```


# Summary

To address the revision comments from the 1st round review, we decided to further consider regional brain measures (e.g., cortical thickness and volume of each of the 34 regions of interest [ROIs] per hemisphere based on Desikan-Killiany Parcellation). 

This rmd file takes the raw datasets and outputs the datasets ready for analyese for the structural brain outcomes (i.e., global GMV, sGMV, WMV, mean CT, regional GMV and regional CT). The output datasets use ABCD release 5.

# Regional brain measure data

## Variable dictionary
```{r, eval = FALSE}
# In the main dataset
# Frontal
ROI_frontal = c(`Super Frontal` = "superiorfrontal",
                `Rostral Middle Frontal` = "rostralmiddlefrontal",
                `Caudal Middle Frontal` = "caudalmiddlefrontal",
                `Pars Opercularis` = "parsopercularis",
                `Pars Triangularis` = "parstriangularis",
                `Pars Orbitalis` = "parsorbitalis",
                `Lateral Orbitofrontal` = "lateralorbitofrontal",
                `Medial Orbitofrontal` = "medialorbitofrontal",
                `Precentral` = "precentral",
                `Paracentral` = "paracentral",
                `Frontal Pole` = "frontalpole",
                `Rostral Anterior` = "rostralanteriorcingulate",
                `Caudal Anterior` = "caudalanteriorcingulate"
                )

# Parietal
ROI_parietal = c(`Superior Parietal` = "superiorparietal",
                 `Inferior Parietal` = "inferiorparietal",
                 `Supramarginal` = "supramarginal",
                 `Postcentral` = "postcentral",
                 `Precuneus` = "precuneus",
                 `Posterior` = "posteriorcingulate",
                 `Isthmus` = "isthmuscingulate"
                 )

# Temporal 
ROI_temporal = c(`Superior Temporal` = "superiortemporal",
                 `Middle Temporal` = "middletemporal",
                 `Inferior Temporal` = "inferiortemporal",
                 `Banks of the Superior Temporal Sulcus` = "bankssts",
                 `Fusiform` = "fusiform",
                 `Transverse Temporal` = "transversetemporal",
                 `Entorhinal` = "entorhinal",
                 `Temporal Pole` = "temporalpole",
                 `Parahippocampal` = "parahippocampal",
                 `Insula` = "insula"
                 )

# Occipital
ROI_occipital = c(`Lateral Occipital` = "lateraloccipital",
                  `Lingual` = "lingual",
                  `Cuneus` = "cuneus",
                  `Pericalcarine` = "pericalcarine"
                  )

# Other?
# ROI_other = c(`Insula` = "insula")

ROI_dict = rbind(cbind(lobe = "Frontal", ROI_label = names(ROI_frontal), var_names = ROI_frontal),
                 cbind(lobe = "Parietal", ROI_label = names(ROI_parietal), var_names = ROI_parietal),
                 cbind(lobe = "Temporal", ROI_label = names(ROI_temporal), var_names = ROI_temporal),
                 cbind(lobe = "Occipital", ROI_label = names(ROI_occipital), var_names = ROI_occipital)
                 # cbind(lobe = "Other", ROI_label = names(ROI_other), var_names = ROI_other)
                 ) %>% as.data.frame

# In the longitudinal ABCD dataset

# variable dictionary for longitudinal ABCD
# Frontal
ROI_frontal = c(`Super Frontal` = "sufr",
                `Rostral Middle Frontal` = "rrmdfr",
                `Caudal Middle Frontal` = "cdmdfr",
                `Pars Opercularis` = "parsopc",
                `Pars Triangularis` = "parstgris",
                `Pars Orbitalis` = "parsobis",
                `Lateral Orbitofrontal` = "lobfr",
                `Medial Orbitofrontal` = "mobfr",
                `Precentral` = "precn",
                `Paracentral` = "paracn",
                `Frontal Pole` = "frpole",
                `Rostral Anterior` = "rracate",
                `Caudal Anterior` = "cdacate"
                )

# Parietal
ROI_parietal = c(`Superior Parietal` = "supl",
                 `Inferior Parietal` = "ifpl",
                 `Supramarginal` = "sm",
                 `Postcentral` = "postcn",
                 `Precuneus` = "pc",
                 `Posterior` = "ptcate",
                 `Isthmus` = "ihcate"
                 )

# Temporal 
ROI_temporal = c(`Superior Temporal` = "sutm",
                 `Middle Temporal` = "mdtm",
                 `Inferior Temporal` = "iftm",
                 `Banks of the Superior Temporal Sulcus` = "bankssts",
                 `Fusiform` = "fusiform",
                 `Transverse Temporal` = "trvtm",
                 `Entorhinal` = "ehinal",
                 `Temporal Pole` = "tmpole",
                 `Parahippocampal` = "parahpal",
                 `Insula` = "insula"
                 )

# Occipital
ROI_occipital = c(`Lateral Occipital` = "locc",
                  `Lingual` = "lingual",
                  `Cuneus` = "cuneus",
                  `Pericalcarine` = "pericc"
                  )

# # Other?
# ROI_other = c()

ROI_dict_long_ABCD = rbind(cbind(lobe = "Frontal", ROI_label = names(ROI_frontal), var_names_ABCD = ROI_frontal),
                           cbind(lobe = "Parietal", ROI_label = names(ROI_parietal), var_names_ABCD = ROI_parietal),
                           cbind(lobe = "Temporal", ROI_label = names(ROI_temporal), var_names_ABCD = ROI_temporal),
                           cbind(lobe = "Occipital", ROI_label = names(ROI_occipital), var_names_ABCD = ROI_occipital)
                           # cbind(lobe = "Other", ROI_label = names(ROI_other), var_names_ABCD = ROI_other)
                           ) %>% as.data.frame

ROI_dict = merge(ROI_dict, ROI_dict_long_ABCD)

write.csv(ROI_dict, "data/analysis_data/ROI/ROI_dict_DKparcellation.csv")
```

```{r}
ROI_dict <- readr::read_csv("data/analysis_data/ROI/ROI_dict_DKparcellation.csv")
```

## Data cleaning

```{r}
library(readr)
# data containing whole brain measures and regional measures
data_ROI <- read_csv("data/raw_data/cn_data_clean_dkmorph_26Jul2023.csv")

var_list = c("participant", "study", "age_years", "sex_01", "dx", "site", "GMV_10000", "sGMV_10000", "WMV_10000",
             # ROI Vol
             paste0("lh_Vol_", ROI_dict$var_names),
             paste0("rh_Vol_", ROI_dict$var_names),
             # CT
             paste0("lh_CT_", ROI_dict$var_names),
             paste0("rh_CT_", ROI_dict$var_names),
             # surface area
             paste0("lh_SA_", ROI_dict$var_names),
             paste0("rh_SA_", ROI_dict$var_names))
```

Same data cleaning as the initial submission was conducted here.

+ study `PPMI` & `FemaleASD`: everyone has the same value in their follow-ups. Only the baseline measurements are remained.

```{r}
# remove duplicate records
data_ROI_clean = subset(data_ROI, !duplicated(data_ROI))

# PPMI:
PPMI = data_ROI_clean %>% filter(study == "PPMI") %>% group_by(study, participant) %>% slice_min(age_years)
data_ROI_clean = rbind(subset(data_ROI_clean, study != "PPMI"), PPMI)
# FemaleASD
FemaleASD = data_ROI_clean %>% filter(study == "FemaleASD") %>% group_by(study, participant) %>% slice_min(age_years)
data_ROI_clean = rbind(subset(data_ROI_clean, study != "FemaleASD"), FemaleASD)

# UCSD
data_ROI_clean = subset(data_ROI_clean, study != "UCSD")
```


## Replace with Longitudinal ABCD (release 5)

```{r}
long_ABCD <- read_csv("data/analysis_data/ABCD_release5.csv") %>% as.data.frame() 
ROI_dict = read_csv("data/analysis_data/ROI/ROI_dict_DKparcellation.csv")
# Volume outcome
## Left
var_vol_LH = paste0("lh_Vol_", ROI_dict$var_names)
## Right
var_vol_RH = paste0("rh_Vol_", ROI_dict$var_names)
# Cortical Thickness
## Left
var_CT_LH = paste0("lh_CT_", ROI_dict$var_names)
## Right
var_CT_RH = paste0("rh_CT_", ROI_dict$var_names)

# Surface area
## Left
var_SA_LH = paste0("lh_SA_", ROI_dict$var_names)
## Right
var_SA_RH = paste0("rh_SA_", ROI_dict$var_names)

long_ABCD = long_ABCD[, c("src_subject_id", "dx", "age_years", "sex_01", "eventname", "site", "GMV", "sGMV", "WMV",
                          var_vol_LH, var_vol_RH, var_CT_LH, var_CT_RH, var_SA_LH, var_SA_RH
                          )]

# SELECTING CN SUBJECTS
long_ABCD = subset(long_ABCD, dx == "CN")

long_ABCD = long_ABCD[order(long_ABCD$src_subject_id, long_ABCD$age_years, decreasing = FALSE), ]

# removing the rows with missing age 
long_ABCD = subset(long_ABCD, !is.na(age_years))


# generate a new "numeric" id
# Note: geeglm has problem with ID values containing characters
long_ABCD %<>% group_by(src_subject_id) %>% mutate(temp_id = cur_group_id()) 


# rescale GMV, sGMV, and WMV
long_ABCD$GMV_10000 = long_ABCD$GMV / 10000
long_ABCD$sGMV_10000 = long_ABCD$sGMV / 10000
long_ABCD$WMV_10000 = long_ABCD$WMV / 10000

# data reformatting
long_ABCD$participant = long_ABCD$temp_id
long_ABCD$study = "ABCD"




# put it back to the full dataset
var_list = c("participant", "study", "age_years", "sex_01", "dx", "site", "GMV_10000", "sGMV_10000", "WMV_10000",
             var_vol_LH, var_vol_RH, var_CT_LH, var_CT_RH, var_SA_LH, var_SA_RH)

data_ROI_clean = data_ROI_clean[, var_list]

# remove the cross-sectional ABCD
data_ROI_clean = subset(data_ROI_clean, study != "ABCD")
# put the longitudinal ABCD back
data_ROI_clean = rbind(data_ROI_clean, long_ABCD[, var_list])



# save the cleaned ABCD release 5
write.csv(long_ABCD[, var_list], "data/analysis_data/ABCD_release5_clean.csv")
```

## Global average CT

The global CT is defined as the average CT across all the DK regions weigthed by the region surface areas.

```{r}
## Left
var_CT_LH = paste0("lh_CT_", ROI_dict$var_names)
## Right
var_CT_RH = paste0("rh_CT_", ROI_dict$var_names)

# Surface area
## Left
var_SA_LH = paste0("lh_SA_", ROI_dict$var_names)
## Right
var_SA_RH = paste0("rh_SA_", ROI_dict$var_names)

data_ROI_clean %<>% as.data.frame()
data_ROI_clean$CT = sapply(1:nrow(data_ROI_clean), 
              function(i){
                x = as.numeric(data_ROI_clean[i, c(var_CT_LH, var_CT_RH)])
                w = as.numeric( data_ROI_clean[i, c(var_SA_LH, var_SA_RH)])
                CT = weighted.mean(x = x, 
                                   w = w)
                return(CT)
              })

```


## Save
```{r}
# if GMV is missing, all ROI measures are missing. 
# only use the complete cases for the ROI outcomes
data_ROI_clean_na_omit = na.omit(data_ROI_clean)

# complete-case data
write.csv(data_ROI_clean_na_omit, "data/analysis_data/ROI/analysis_data_ROI_complete_case.csv")
# not complete-case data
write.csv(data_ROI_clean, "data/analysis_data/ROI/analysis_data_ROI.csv")
```


**Compare with the data used for initial submission**:

```{r}
# data used for initial submission
analysis_data <- read_csv("data/analysis_data/whole_brain_measure/old_initial_submission/analysis_data_combat.csv")
analysis_data %<>% as.data.frame

# data that is going to be used for the 1st revision
data_ROI_clean = read_csv("data/analysis_data/ROI/analysis_data_ROI_complete_case.csv")

data_new = cbind(study = data_ROI_clean$study, R = "New")
data_old = cbind(study = analysis_data$study, R = "Previous")

data_combine = rbind(data_new, data_old) %>% as.data.frame()
table1(~ study|R, data = data_combine)

```

## Study feature summary table
```{r}
# new study summary table 

# FUNCTION for creating study summary table
sum_tab = function(data) {
  # independent data
  ind_data = data %>% group_by(study, participant) %>% slice_sample(n = 1) %>% ungroup()
  # Note: there may be some subjects from different study but having same subject IDs.
  tab1 = ind_data %>% group_by(study) %>% summarise(N = length(unique(participant)))
  tab2 = data %>% group_by(study) %>% summarise(tot_obs = length(participant) )
  tab = merge(tab1, tab2)
  tab = tab[order(tab$N, decreasing = TRUE),]
  rownames(tab) = as.character(1:nrow(tab))
  tab %<>% mutate(Design = ifelse(N == tot_obs, "Cross-sectional", "Longitudinal"))
  
  features = data %>% group_by(study) %>% summarise(med_age = median(age_years), 
                                                     mean_age = moment(age_years), 
                                                     m2_age = moment(age_years, order = 2, center = TRUE) %>% sqrt(), # (2nd central moment)^(1/2)
                                                     m3_age = moment(age_years, order = 3, center = TRUE) %>% nthroot(3), # (3rd central moment)^(1/3)
                                                     mean_sex = moment(sex_01),
                                                     m2_sex = moment(sex_01, order = 2, center = TRUE) %>% sqrt(),
                                                     m3_sex = moment(sex_01, order = 3, center = TRUE) %>% nthroot(3)
                                                     )
  
  # proportion of records that have at least one of ROI variables has missing value
  tab = merge(tab, features)
  
  
  return(tab)
}


data_ROI_clean = read_csv("data/analysis_data/ROI/analysis_data_ROI_complete_case.csv")


study_info_clean = sum_tab(data_ROI_clean) 

write.csv(study_info_clean, "data/analysis_data/ROI/study_info_clean_ROI.csv")

```

## Removing site effects

```{r}
# Note: different subjects may have missing values in different variables. For Combat, we need to remove the missing values in advance. 
# so for different ROI measures, we may need to remove different records.

library(neuroCombat)
library(longCombat)

study_info_clean = read_csv("data/analysis_data/ROI/study_info_clean_ROI.csv")
data_ROI_clean = read_csv("data/analysis_data/ROI/analysis_data_ROI_complete_case.csv") %>% as.data.frame
ROI_dict <- readr::read_csv("data/analysis_data/ROI/ROI_dict_DKparcellation.csv") %>% as.data.frame

ROI_var_list = c(# ROI Vol
                 paste0("lh_Vol_", ROI_dict$var_names),
                 paste0("rh_Vol_", ROI_dict$var_names),
                 # CT
                 paste0("lh_CT_", ROI_dict$var_names),
                 paste0("rh_CT_", ROI_dict$var_names))


combat_data = NULL
for (i in 1:nrow(study_info_clean)){
  
  study_i = study_info_clean$study[i]
  design_i = study_info_clean$Design[i]
  cat("Study:", as.character(study_i), "\n")
  cat("    Design:", design_i, "\n")
  # data
  data_i = subset(data_ROI_clean, study == study_i)
  data_i = data_i[order(data_i$participant, data_i$age_years),] # sort
  ## generating a unique record indicator for merging combat outcome back
  ## Some subjects may have two visits that are at the same age (i.e., the visit interval is too short).
  # data_i$U_id = 1:nrow(data_i)
  
  if (length(unique(data_i$site)) == 1) {
    cat("only 1 site in this dataset, skip. \n")
    # the combat variable = the original variable
    data_i[, paste0(ROI_var_list, "_combat")] = data_i[, ROI_var_list] 
    data_i = data_i[, c("study", "participant", "age_years", "sex_01", "site", paste0(rep(ROI_var_list, each = 2), c("", "_combat")))]
  } else {
      if (design_i == "Longitudinal") {
        
        obj_combat = longCombat(idvar = "participant", 
                               timevar = "age_years", 
                               batchvar = "site", 
                               features = c(ROI_var_list),
                               formula = "ns(age_years, df = 2) + sex_01",
                               ranef = "(1|participant)",
                               data = data_i,
                               verbose = FALSE)
        
        data_i[, paste0(ROI_var_list, "_combat")] = obj_combat$data_combat[, paste0(ROI_var_list, ".combat")]
        
        data_i = data_i[order(data_i$participant, data_i$age_years), ]
        
      # reorder the columns
      data_i = data_i[, c("study", "participant", "age_years", "sex_01", "site", paste0(rep(ROI_var_list, each = 2), c("", "_combat")))]
    } # end of `if (design_i == "longitudinal")`  
    
    if (design_i == "Cross-sectional"){

      obj_combat = neuroCombat(dat = data_i[, ROI_var_list] %>% t(), mod = model.matrix(~ ns(age_years, df = 2) + sex_01, data=data_i),
                             batch = data_i$site, eb = TRUE, verbose = FALSE)

      
      data_i[, paste0(ROI_var_list, "_combat")] = t(obj_combat$dat.combat)
      
      # reorder the columns
      data_i = data_i[, c("study", "participant", "age_years", "sex_01", "site", paste0(rep(ROI_var_list, each = 2), c("", "_combat")))]
    }
      
  }
    combat_data = rbind(combat_data, data_i)
}


# write.csv(combat_data, "data/analysis_data/ROI/analysis_data_ROI_complete_case_combat.csv")

```

# Whole Brain Measures


```{r}
# the data used for initial submission
analysis_data_ <- read_csv("data/analysis_data/whole_brain_measure/old_initial_submission/analysis_data_combat.csv")

# ABCD release 5
ABCD_release5_clean <- read_csv("data/analysis_data/ABCD_release5_clean.csv")

var_list = c("participant", "study", "age_years", "sex_01", "dx", "site", "GMV_10000", "sGMV_10000", "WMV_10000")

ABCD = ABCD_release5_clean[, var_list]

# remove rows containing missing values
ABCD = na.omit(ABCD)

# longCombat
ABCD_combat = longCombat(idvar = "participant", 
                       timevar = "age_years", 
                       batchvar = "site", 
                       features = c("GMV_10000", "sGMV_10000", "WMV_10000"),
                       formula = "ns(age_years, df = 2) + sex_01",
                       ranef = "(1|participant)",
                       data = ABCD,
                       verbose = FALSE)
        
ABCD[, paste0(c("GMV_10000", "sGMV_10000", "WMV_10000"), "_combat")] = ABCD_combat$data_combat[, paste0(c("GMV_10000", "sGMV_10000", "WMV_10000"), ".combat")]
        
ABCD = ABCD[order(ABCD$participant, ABCD$age_years), ]

# replace with ABCD release 5 after longCombat
var_list = c("participant", "study", "age_years", "sex_01", "dx", "site", "GMV_10000", "sGMV_10000", "WMV_10000", paste0(c("GMV_10000", "sGMV_10000", "WMV_10000"), "_combat"))

analysis_data_combat = subset(analysis_data_combat, !study == "ABCD")
analysis_data_combat = rbind(analysis_data_combat[, var_list], ABCD[, var_list])

write.csv(analysis_data_combat, "data/analysis_data/whole_brain_measure/new/analysis_data_whole_brain_combat.csv")

# study info summary table
study_info_whole_brain = sum_tab(analysis_data_combat) 
write.csv(study_info_whole_brain, "data/analysis_data/whole_brain_measure/new/study_info_whole_brain.csv")
```


**Comparing the new and old data**:

```{r}
old =  readr::read_csv("data/analysis_data/whole_brain_measure/old_initial_submission/analysis_data_combat.csv")
old$version = "old"

analysis_data = readr::read_csv("data/analysis_data/whole_brain_measure/new/analysis_data_whole_brain_combat.csv")
analysis_data$version = "new"
temp = rbind(old[, c("participant", "study", "version")], analysis_data[, c("participant", "study", "version")])
table1::table1(~ study | version, data = temp)

# number of subjects from each study
new_1 = analysis_data %>% group_by(study, participant) %>% slice_head(n = 1)
old_1 = old %>% group_by(study, participant) %>% slice_head(n = 1)

temp = rbind(old_1[, c("participant", "study", "version")], new_1[, c("participant", "study", "version")])
table1::table1(~ study | version, data = temp)
# paste0(old$study, old$participant) %>% unique %>% length
# old$participant %>% unique %>% length
# 
# paste0(analysis_data$study, analysis_data$participant) %>% unique %>% length
# 
# old %>% apply(2, FUN = function(x){sum(is.na(x))})
```



## Global CT
```{r}
data_CT = read_csv("data/analysis_data/ROI/analysis_data_ROI_complete_case.csv")
data_CT = data_CT[, c("participant", "study", "age_years", "sex_01", "dx", "site","GMV_10000", "CT")]

# Combat / Long combat

library(neuroCombat)
library(longCombat)

study_info_clean = read_csv("data/analysis_data/ROI/study_info_clean_ROI.csv")

combat_data = NULL
for (i in 1:nrow(study_info_clean)){
  
  study_i = study_info_clean$study[i]
  design_i = study_info_clean$Design[i]
  cat("Study:", as.character(study_i), "\n")
  cat("    Design:", design_i, "\n")
  # data
  data_i = subset(data_CT, study == study_i)
  data_i = data_i[order(data_i$participant, data_i$age_years),] # sort
  ## generating a unique record indicator for merging combat outcome back
  ## Some subjects may have two visits that are at the same age (i.e., the visit interval is too short).
  # data_i$U_id = 1:nrow(data_i)
  
  if (length(unique(data_i$site)) == 1) {
    cat("only 1 site in this dataset, skip. \n")
    # the combat variable = the original variable
    data_i[, "CT_combat"] = data_i[, "CT"] 
    data_i = data_i[, c("study", "participant", "age_years", "sex_01", "site", "CT", "CT_combat")]
  } else {
      if (design_i == "Longitudinal") {
        
        obj_combat = longCombat(idvar = "participant", 
                               timevar = "age_years", 
                               batchvar = "site", 
                               features = c("GMV_10000", "CT"),
                               formula = "ns(age_years, df = 2) + sex_01",
                               ranef = "(1|participant)",
                               data = data_i,
                               verbose = FALSE)
        
        data_i[, "CT_combat"] = obj_combat$data_combat[, "CT.combat"]
        
        data_i = data_i[order(data_i$participant, data_i$age_years), ]
        
      # reorder the columns
      data_i = data_i[, c("study", "participant", "age_years", "sex_01", "site", "CT", "CT_combat")]
    } # end of `if (design_i == "longitudinal")`  
    
    if (design_i == "Cross-sectional"){

      obj_combat = neuroCombat(dat = data_i[, c("GMV_10000", "CT")] %>% t(), mod = model.matrix(~ ns(age_years, df = 2) + sex_01, data=data_i),
                             batch = data_i$site, eb = TRUE, verbose = FALSE)
      
      temp =  t(obj_combat$dat.combat)
      data_i[, "CT_combat"] = temp[, "CT"]
      
      # reorder the columns
      data_i = data_i[, c("study", "participant", "age_years", "sex_01", "site", "CT", "CT_combat")]
    }
      
  }
    combat_data = rbind(combat_data, data_i)
}

write.csv(combat_data, "data/analysis_data/whole_brain_measure/new/CT/analysis_data_global_CT_combat.csv")
```


```{r}
CT_study_info_clean = sum_tab(data_CT) 

write.csv(CT_study_info_clean, "data/analysis_data/whole_brain_measure/new/CT/study_info_clean_global_CT.csv")
```




